---
title: "Abalones"
author: "Melinda Higgins"
date: "January 23, 2019"
output: html_document
---

```{r setup, include=FALSE}
# setup knitr options
knitr::opts_chunk$set(echo = TRUE)

# load packages needed
library(readr)
library(dplyr)
library(ggplot2)

# Load the abalone dataset and provide variable names
abalone <- readr::read_csv(file = "https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data",
                           col_names = FALSE)

names(abalone) <- c("sex","length",
                    "diameter","height",
                    "wholeWeight","shuckedWeight",
                    "visceraWeight","shellWeight","rings")
```

## Abalones Dataset from UCI Repository

The `abalone` dataset is available from the UCI data repository at [https://archive.ics.uci.edu/ml/datasets/abalone](https://archive.ics.uci.edu/ml/datasets/abalone).

## Variables in Abalone Dataset

The variables in the abalone dataset are listed below.

```{r}
names(abalone)
```

## Summary statistics of variables in `abalone`

```{r}
summary(abalone)
```

## Specific statistics within text

```{r, echo = FALSE}
# this code chunk is not shown in the final document
# but we use this code chunk to save a result
# and use it below - the mean of height
# assign the mean height to object meanHt
meanHt <- abalone %>% 
  pull(height) %>%
  mean()
```

We can use `rmarkdown` with `R` to embed `R` code within text to show the result in the final document instead of the code.

For example, the average height of the `abalone` is `r meanHt`.

## Histogram of `abalone` heights

When typing text in `rmarkdown` we can add formatting like making works **BOLD** or adding other emphasis using _italics_.

We can also add bullets:

* What do you notice about the abalone heights?
* What could we do to investigate this issue further?

```{r}
# make a histogram of height
ggplot(data = abalone, aes(x = height)) +
  geom_histogram()
```

## Scatterplot of `abalone` `rings` by `shellWeight`

```{r}
# make a scatterplot of rings by shellWeight
# color points by sex
ggplot(data = abalone, 
       aes(x = shellWeight, y = rings)) +
  geom_point(aes(color = sex))
```

