---
title: "NRSG 741 - Homework 2 - ANSWER KEY"
author: "Melinda Higgins"
date: "03/25/2019"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

## INSTRUCTIONS

* Use this Rmarkdown file `N741Homework02.Rmd` to get started.
* Change the author to YOUR NAME
* Note: This Rmarkdown file has one R code chunk at the top that reads in the dataset and loads the R packages you will need.
* After each question below, insert an R code chunk to enter the R code needed to answer that question. Do this for each question.
* Outside of the R code chunk, type in any text needed to provide explanation or answer the questions further.

```{r setup, include=FALSE}
# set up chunk output options
# set to FALSE to clean up final output
# leave TRUE to help with debugging initially
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = TRUE)
knitr::opts_chunk$set(message = TRUE)
knitr::opts_chunk$set(error = TRUE)

# Get data
# learn more at https://archive.ics.uci.edu/ml/datasets/abalone

# Load the abalone dataset using read_csv() function
# from readr package
library(readr)
abalone <- readr::read_csv(file = "https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data",
                           col_names = FALSE)

# put in the variables names
# see details at https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.names
names(abalone) <- c("sex","length",
                    "diameter","height",
                    "wholeWeight","shuckedWeight",
                    "visceraWeight","shellWeight","rings")

# Load dplyr and ggplot2 packages
library(dplyr)
library(ggplot2)
```

## *Due Date* is 13 February 2019

This homework is meant to further your `dplyr` and `ggplot2` skills. 

## Abalones Dataset from UCI Repository

For this homework, you will keep working with the `abalone` dataset from the UCI data repository at [https://archive.ics.uci.edu/ml/datasets/abalone](https://archive.ics.uci.edu/ml/datasets/abalone).

Use tools within the `dplyr` package as much as possible to answer the following questions.

##### **Question 1:** What kind of R object is the `abalone` dataset?

**ANSWER KEY**  

You could use either the `class()` or `str()` functions to give you details about the `abalone` dataset object.

```{r}
# insert R code here to answer question 1

class(abalone)

str(abalone)
```

**ANSWER KEY**

The `abalone` dataset is read in as a `data.frame`, using the `read_csv()` function from the `readr` package which is part of the `tidyverse`, actually makes it a `"spec_tbl_df" "tbl_df"      "tbl"         "data.frame"` - a tibble data frame. You can learn more at [https://www.tidyverse.org/articles/2018/12/readr-1-3-1/](https://www.tidyverse.org/articles/2018/12/readr-1-3-1/).

##### **Question 2:** How many observations are in the `abalone` dataset?

**ANSWER KEY**

To answer this question, you can use either the `str()` or `dim()` functions.

```{r}
dim(abalone)
```

**ANSWER KEY**

Based on either the `str()` output shown above or the `dim()` results, there are `r dim(abalone)[1]` observations in the `abalone` dataset.

##### **Question 3:** For shucked weight, how many abalones weigh more than 0.8 grams?

**ANSWER KEY**

Using the `filter()` function from the `dplyr` package is useful for extracting cases (observations or rows) that meet the specified criteria defined inside the `filter()` function. Only rows for which the filter is TRUE are retained.

```{r}
abalone %>%
  filter(shuckedWeight > 0.8) %>%
  dim()
```

There are 148 abalones with a shucked weight > 0.8 grams.

##### **Question 4:** How many abalones have shucked weights larger than their whole weight? 

_(HINT: create a new variable using mutate and then filter)_

**ANSWER KEY**

This problem could have been solved using either `mutate()` or the `filter() ` function. Both approached should yield the same answer.

```{r}
abalone %>%
  mutate(shuckedHigh = shuckedWeight > wholeWeight) %>%
  filter(shuckedHigh == TRUE) %>%
  dim()

# alternate approach without mutate

abalone %>%
  filter(shuckedWeight > wholeWeight) %>%
  dim()
```

There are 4 abalones with shucked weight greater than their whole weight which should not be correct. These abalones have measurement errors.

---

Create a subset containing only infants `sex == "I"`

##### **Question 5:** How many infants are in this subset?

**ANSWER KEY**

Create the subset first and then find the dimensions to get number of rows.

```{r}
# Create subset
abaloneI <- abalone %>%
  filter(sex == "I")

# Find dimensions
dim(abaloneI)
```

There are `r dim(abaloneI)[1]` infant abalones in this dataset.

---

Show off your `dplyr` skills with `group_by()`

##### **Question 6:** What is the average whole weight for each abalone sex (get whole weight means for females "F", males "M" and infants "I" separately)?

**ANSWER KEY**

You can use either `summarise()` or `summarise_all()` functions from `dplyr` package.

```{r}
abalone %>%
  group_by(sex) %>%
  summarise(meanwt = mean(wholeWeight))

abalone %>%
  group_by(sex) %>%
  select(wholeWeight) %>%
  summarise_all(mean)
```

##### **Question 7:** Get the means for the abalone length and height by sex?

**ANSWER KEY**

You can use either `summarise()` or `summarise_all()` functions from `dplyr` package. This is very similar to problem above, notice only the variable names get updated.

```{r}
abalone %>%
  group_by(sex) %>%
  summarise(meanlt = mean(length),
            meanht = mean(height))

abalone %>%
  group_by(sex) %>%
  select(length, height) %>%
  summarise_all(mean)
```


----

## Test your graphing skills using `ggplot2`

Using the `abalone` dataset, create the following graphics/figures using `ggplot()` and associated `geom_xxx()` functions.

##### **Question 8:** Create a histogram of abalone whole weight 

_What do you notice about the distribution (any outliers or skewness)?_

**ANSWER KEY**

You want to use the `geom_histogram()` function from the `ggplot2` package. In the initial `ggplot()` step, you only have to define one aesthetic (`aes`) for `wholeWeight`.

```{r}
# simple histogram
ggplot(abalone, aes(x=wholeWeight)) +
  geom_histogram()

# optional - add colors and labels and a title
ggplot(abalone, aes(x=wholeWeight)) +
  geom_histogram(color = "black",
                 fill = "light blue") +
  xlab("Whole Weight in grams") + 
  ylab("Frequencies") + 
  ggtitle("Histogram of Abalone Whole Weights")
```

There are a few large abalones with weights above 2.5 grams.

##### **Question 9:** Create side-by-side boxplots of the number of rings by gender 
_HINT use `geom_boxplot` with x = sex and y = rings_ 

**ANSWER KEY**

Use similar approach to above, but now you have two aesthetics (`aes`) instead of just one for the histogram above. The two aesthetics (`aes`) are `sex` and `rings`. You need `geom_boxplot()` to draw the boxplots.

```{r}
ggplot(abalone, aes(x=sex, y=rings)) +
  geom_boxplot()
```


##### **Question 10:** Create a scatterplot of the whole weight on the X axis and shucked weight on the Y axis and color the points by sex

**ANSWER KEY**

Scatterplots also need two aesthetics (`aes`) - in this case are `wholeWeight` for "x" and `shuckedWeight` for "y". You also need `geom_point()` and color the points by `sex` using `aes(color)` inside `geom_point()`.

```{r}
ggplot(abalone, aes(x=wholeWeight, y=shuckedWeight)) +
  geom_point(aes(color = sex))
```

